Topological Equivariant Artist Model

### 2021-03-16 paper/paper.tex
abstract

\usepackage{microtype}                 % use micro-typography (slightly more compact, better to read)
\PassOptionsToPackage{warn}{textcomp}  % to address font issues with \textrightarrow
\usepackage{textcomp}                  % use better special symbols
\usepackage{mathptmx}                  % use matching math font
\usepackage{times}                     % we use Times as the main font
\renewcommand*\ttdefault{txtt}         % a nicer typewriter font
\usepackage{cite}                      % needed to automatically sort the references
%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.
\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
\usepackage{mathrsfs} % math script fonts
\usepackage{tabulary}
\usepackage{subfig}
\usepackage{multicol}
\usepackage{tikz}
\usetikzlibrary{cd} % commutative diagrams
\usepackage[switch]{lineno}
\usepackage{minted}
\setminted[python]{fontsize=\scriptsize, 
                   linenos,
                   numbersep=8pt,
                   autogobble, 
                   frame=lines,
                   framesep=3mm} 
\usepackage{notation} % move this later

@@

%% indicate IEEE Member or Student Member in form indicated below
\author{Hannah Aizenman, Thomas Caswell, Michael Grossberg}
\authorfooter{
%% insert punctuation at end of each item
\item
 Hannah Aizenman and Michael Grossberg are with the Computer Science department, City College of New York. E-mail: roy.g.biv@aol.com.
\item
 Thomas Caswell is with Brookhaven National Lab E-mail: ed.grimley@aol.com.
}

@@

%% Abstract section.
\abstract{ A critical aspect of data visualization is that the graphical representation of data is expected to match the properties of the data; this fails when order is not preserved in representations of ordinal data or scale for numerical data. In this work, we propose that the mathematical notions of equivariance and topology formalizes the expectation of matching properties. We developed a model we call the topological artist model (TAM) in which data and graphics can be viewed as sections of fiber bundles. This model allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that translates the dataset connectivity into graphical elements. Furthermore, our model supports an algebraic sum operation such that more complex visualizations can be built from simple ones. We illustrate the application of the model through case studies of a scatter plot, line plot, and heatmap. We show that this model can be implemented with a small prototype.} % end of abstract

@@

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}
\maketitle
The aim of this work is to rearchitecture Matplotlib to take advantage of developments in software design, data structures, and visualization to improve consistency, reusability, and discoverability, so domain specific tool developers can build structure preserving visualization tools. The contribution of this work is (a) topology preserving relationship between data and graphic via continuous maps, (b) property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions, (c) functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model, (d) prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model.

@@

Specifically, \vartist is the equivariant map from $\dtotal^{\prime}$ to a specific graphic $\gsection \in \Gamma(H)$ 
\begin{equation}
    \label{eq:math:artist:diagram}
    \begin{tikzcd}
        \dtotal^{\prime} \arrow[r, "\vchannel"] \arrow[rd, "\pi"'] & \vtotal \arrow[d, "\pi"] & \vindex^*\vtotal \arrow[r, "\vmark"] \arrow[d, "\vindex^*\pi"'] \arrow[l, "\vindex^*"'] & \gtotal \arrow[ld, "\pi"] \\
                                              & \dbase                  & \gbase \arrow[l, "\vindex"']                                              &                    
        \end{tikzcd}
\end{equation}
where the input can be point wise $\dsection(\dbasepoint)\mid \dbasepoint \in \dbase$. The visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ is the latent space of possible parameters of a visualization type, such as a scatter or line plot. As with the data and graphic bundles, the visual bundle is  defined by the projection map $\pi$
\begin{equation}
    \begin{tikzcd}[ampersand replacement=\&]
        \vfiber \arrow[r, hook] \& \vtotal \arrow[d, "\pi"'] \\
                          \& \dbase \arrow[u, "\vsection"', bend right]
    \end{tikzcd}
\end{equation}
where \vsection\ is the visual variable encoding\cite{bertinSemiologyGraphicsDiagrams2011a} of the data section \dsection. The visual fiber \vfiber\ is defined in terms of the input parameters of the visualization library's plotting functions; by making these parameters explicit components of the fiber, we can build consistent definitions and expectations of how these parameters behave.

@@

\subsubsection{Visualization Assembly}
\begin{figure}[htb]
  \centering
  \includegraphics[width=\columnwidth]{path_of_q}
  \caption{$\vchannel_i$ functions convert data $\dsection_i$ to visual characteristics $\vsection_i$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection$ such that there is a map \vindex\ preserving the continuity of the data.} 
  \label{fig:math:artist:q}
\end{figure}
As shown in \autoref{fig:math:artist:q}, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings \vsection\, reduce the encodings into a graphic \gsection. While it may seem intuitive that visualizations that generate the same graphic should do so consistently given the same input, we formalize this constraint such that it can be specified as a constraint of \vmark. We define the equivariant map as  $\vmark: \vsection \mapsto \gsection$ and an action on the subset of graphics $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ that \vmark\ can generate. We then define the constraint on \vmark such that if \vmark\ is applied to two visual sections $\vsection,\vsection^{\prime}$ that generate the same \gsection\, then the output of $\vsection,\vsection^{\prime}$ acted on by the same monoid $m$ must be the same.

@@

In \autoref{fig:teaser}, we illustrate the output of a minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and an image. The scatter plot can be defined as $\vmark(xpos, ypos)(\alpha, \beta)$ where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase. The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in \autoref{fig:math:artist:base}
\begin{align*}
x &= size *\alpha \cos(\beta) + xpos \\
y &= size *\alpha \sin(\beta) + ypos
\end{align*}
such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black. In contrast, the line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ in \autoref{fig:teaser} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase\. As shown in \ref{fig:math:arist:base} line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness. The magnitude of the slope is $\lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2}$
such that the normal is  $\hat{n_{1}} = \frac{n_1}{\lvert n \rvert}, \; \hat{n_{2}} = \frac{n_2}{\lvert n \rvert}$ which yields components of \gsection\
\begin{align*}
 x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n_1}(\vindex(\alpha)) \\
 y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n_2}(\vindex(\alpha)) 
\end{align*}
where (x,y) look up the position $\vindex(\alpha)$ on the data and the derivatives $\hat{n_1}, \hat{n_2}$ . The derivatives are then multiplied by a $width$ parameter to specify the thickness.

@@

As shown in \autoref{fig:math:artist:qhat}, given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is a direct map $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$  from \vsection\ over \dbasepoint\ to the section \vsectionpull\ over \gbasepoint. This means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint such that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$. In \autoref{fig:math:artist:qhat} each dot on \vfiber is equivalent to the line on \vfiberpull.

@@

where \mintinline{python}{lambda x: x} is an identity \vchannel\, \mintinline{python}|{'name':None}| maps into \vfiber\ without corresponding \dsection\ to set a constant visual value, and \mintinline{python}{color.Categorical} is a custom \vchannel\ implemented as a class for reusability.  A test for equivariance can be implemented trivially
\begin{minted}{python}
def test_nominal(values, encoder):
    m1 = list(zip(values, encoder(values)))
    random.shuffle(values)
    m2 = list(zip(values, encoder(values)))
    assert sorted(m1) == sorted(m2)
\end{minted}
but is currently factored out of the artist for clarity.

>>>>

\usepackage{microtype}                 % use micro-typography (slightly more compact, better to read)
\PassOptionsToPackage{warn}{textcomp}  % to address font issues with \textrightarrow
\usepackage{textcomp}                  % use better special symbols
\usepackage{mathptmx}                  % use matching math font
\usepackage{times}                     % we use Times as the main font
\renewcommand*\ttdefault{txtt}         % a nicer typewriter font
\usepackage{cite}                      % needed to automatically sort the references
%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.
\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
\usepackage{mathrsfs} % math script fonts
\usepackage{tabulary}
\usepackage{subfig} %ieee does not like subfigure
\usepackage{multicol}
\usepackage{tikz}
\usetikzlibrary{cd} % commutative diagrams
\usepackage[switch]{lineno}
\usepackage{minted}
\setminted[python]{fontsize=\scriptsize, 
                   linenos,
                   numbersep=8pt,
                   autogobble, 
                   frame=lines,
                   framesep=3mm} 
\usepackage{notation} % move this later

@@

%% indicate IEEE Member or Student Member in form indicated below
\author{Hannah Aizenman, Thomas Caswell, Michael Grossberg}
\authorfooter{
%% insert punctuation at end of each item
\item
 Hannah Aizenman and Michael Grossberg are with the Computer Science department, City College of New York. E-mail: haizenman@ccny.cuny.edu, mgrossberg@ccny.cuny.edu.
\item
 Thomas Caswell is with National Synchrotron Light Source II, Brookhaven National Lab E-mail: tcaswell@bnl.gov.
}

@@

%% Abstract section.
\abstract{ A critical aspect of data visualization is that the graphical representation of data is expected to match the structure of the data; this fails, for example, when order is not preserved in representations of ordinal data or proportional distance for numerical data or when discrete data is plotted on a line plot or interpolated as an image. In this work, we propose the topological equivariant artist model (TEAM), wherein we formalize the expectations of attributes and continuity matching using the mathematical notion of equivariance and by modeling data and graphics as fiber bundles. Specifically, our model introduces measurement scale preserving maps on the fiber that transform data components (variables) into visual channels and a continuity preserving map on the base space that transforms dataset connectivity into graphical elements. Furthermore, TEAM has a sum operator to allow for building composite visualizations. We illustrate the application of the model through case studies of a scatter plot, line plot, and heatmap and show that this model can be implemented with a small prototype.}

% end of abstract

@@

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}
\maketitle
The aim of this work is to rearchitecture Matplotlib to take advantage of developments in software design, data structures, and visualization to improve consistency, reusability, and discoverability, so domain specific tool developers can build structure preserving visualization tools. The contribution of this work is 
\begin{enumerate}
  \item topology preserving relationship between data and graphic via continuous maps
  \item property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
  \item functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model.
\end{enumerate}

@@

Specifically, \vartist\ is the equivariant map from $\dtotal^{\prime}$ to a specific graphic $\gsection \in \Gamma(H)$ 
\begin{equation}
    \label{eq:math:artist:diagram}
    \begin{tikzcd}
        \dtotal^{\prime} \arrow[r, "\vchannel"] \arrow[rd, "\pi"'] & \vtotal \arrow[d, "\pi"] & \vindex^*\vtotal \arrow[r, "\vmark"] \arrow[d, "\vindex^*\pi"'] \arrow[l, "\vindex^*"'] & \gtotal \arrow[ld, "\pi"] \\
                                              & \dbase                  & \gbase \arrow[l, "\vindex"']                                              &                    
        \end{tikzcd}
\end{equation}
where the input can be point wise $\dsection(\dbasepoint)\mid \dbasepoint \in \dbase$. The visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ is the latent space of possible parameters of a visualization type, such as a scatter or line plot. As with the data and graphic bundles, the visual bundle is  defined by the projection map $\pi$
\begin{equation}
    \begin{tikzcd}[ampersand replacement=\&]
        \vfiber \arrow[r, hook] \& \vtotal \arrow[d, "\pi"'] \\
                          \& \dbase \arrow[u, "\vsection"', bend right]
    \end{tikzcd}
\end{equation}
where \vsection\ is the visual variable encoding\cite{bertinSemiologyGraphicsDiagrams2011a} of the data section \dsection. The visual fiber \vfiber\ is defined in terms of the input parameters of the visualization library's plotting functions; by making these parameters explicit components of the fiber, we can build consistent definitions and expectations of how these parameters behave.

@@

\subsubsection{Visualization Assembly}
\begin{figure}[htb]
  \centering
  \includegraphics[width=\columnwidth]{path_of_q}
  \caption{$\vchannel_i$ functions convert data $\dsection_i$ to visual characteristics $\vsection_i$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection$ such that there is a map \vindex\ preserving the continuity of the data.} 
  \label{fig:math:artist:q}
\end{figure}
As shown in \autoref{fig:math:artist:q}, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings \vsection\ reduce the encodings into a graphic \gsection. While it may seem intuitive that visualizations that generate the same graphic should do so consistently given the same input, we formalize this constraint such that it can be specified as a constraint of \vmark. We define the equivariant map as  $\vmark: \vsection \mapsto \gsection$ and an action on the subset of graphics $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ that \vmark\ can generate. We then define the constraint on \vmark such that if \vmark\ is applied to two visual sections $\vsection,\vsection^{\prime}$ that generate the same \gsection\, then the output of $\vsection,\vsection^{\prime}$ acted on by the same monoid $m$ must be the same.

@@

In \autoref{fig:teaser}, we illustrate the output of a minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and an image. The scatter plot can be defined as $\vmark(xpos, ypos)(\alpha, \beta)$ where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase. The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in \autoref{fig:math:artist:base}
\begin{align*}
x &= size *\alpha \cos(\beta) + xpos \\
y &= size *\alpha \sin(\beta) + ypos
\end{align*}
such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black. In contrast, the line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ in \autoref{fig:teaser} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase. As shown in \ref{fig:math:arist:base} line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness. The magnitude of the slope is $\lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2}$
such that the normal is  $\hat{n_{1}} = \frac{n_1}{\lvert n \rvert}, \; \hat{n_{2}} = \frac{n_2}{\lvert n \rvert}$ which yields components of \gsection\
\begin{align*}
 x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n_1}(\vindex(\alpha)) \\
 y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n_2}(\vindex(\alpha)) 
\end{align*}
where (x,y) look up the position $\vindex(\alpha)$ on the data and the derivatives $\hat{n_1}, \hat{n_2}$ . The derivatives are then multiplied by a $width$ parameter to specify the thickness.

@@

As shown in \autoref{fig:math:artist:qhat}, given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is a direct map $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$  from \vsection\ over \dbasepoint\ to the section \vsectionpull\ over \gbasepoint. This means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint such that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$. In \autoref{fig:math:artist:qhat} each dot on \vfiber\ is equivalent to the line on \vfiberpull.

@@

where \mintinline{python}{lambda x: x} is an identity \vchannel, \mintinline{python}|{'name':None}| maps into \vfiber\ without corresponding \dsection\ to set a constant visual value, and \mintinline{python}{color.Categorical} is a custom \vchannel\ implemented as a class for reusability.  A test for equivariance can be implemented trivially
\begin{minted}{python}
def test_nominal(values, encoder):
    m1 = list(zip(values, encoder(values)))
    random.shuffle(values)
    m2 = list(zip(values, encoder(values)))
    assert sorted(m1) == sorted(m2)
\end{minted}
but is currently factored out of the artist for clarity.
